A Classical Density-Functional Theory for Describing Water Interfaces

### 2013-08-07 papers/water-saft/paper.tex
Some small changes to water paper Tested everything on node25.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
We develop an improved version of our old functional.
\end{abstract}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Water, the universal solvent, is of critical practical importance, and
a continuum description of water is in high demand for a solvation
model.  A number of recent attempts to develop improved solvation
models for water have built on the approach of classical
density-functional theory (DFT)\cite{jeanmairet2013molecular,
  zhao2011molecular, zhao2011new, ramirez2005direct,
  ramirez2005density, levesque2012solvation, levesque2012scalar}
~\fixme{add citations here, particularly Jeanmairet 2013 and his
  citations 35-44, possibly cutting non-acqueous papers from that
  list}.  Classical DFT is based on a description of a fluid written
as a free energy functional of the density distribution.  There are
two approaches used to construct a classical DFT for water.  The first
is to choose a functional form that is fit to properties of the bulk
liquid at a given temperature and pressure
\cite{jeanmairet2013molecular} ~\fixme{cite Jeanmairet and similar}.
Using this approach, it is possible to construct a functional that
reproduces the exact second-order response function of the liquid
under the fitted conditions.  This class of functional, however, will
be less accurate at other temperatures or pressures---and in the
inhomogeneous scenarios in which solvation models are applied.  The
second approach is construct a functional by applying liquid-state
theory to a model system, and then fit the model to experimental data
such as the equation of state.

%% \fixme{stuff about water}.  An important aspects of water
%% is its ability to form hydrogen bonds.  Each water molecule contains
%% two hydrogen bond donor sites and two acceptor sites.
%% %
%% One promising approach for an efficient continuum description of water
%% is that of classical density-functional theory (DFT), which is an
%% approach for evaluating the free energy and thermally averaged density
%% of fluids in an arbitrary external potential~\cite{ebner1976density}.
%% The foundation of classical DFT is the Mermin
%% theorem\cite{mermin1965thermal}, which extends the Hohenberg-Kohn
%% theorem\cite{hohenberg1964inhomogeneous} to non-zero temperature.
%% Classical DFT is a natural framework for creating a more flexible
%% theory of hydrophobicity that can readily describe interaction of
%% water with arbitrary external potentials---such as potentials
%% describing strong interactions with solutes or surfaces.

A widely used family of models used in the development of classical
density functionals is based on Statistical Associating Fluid Theory
(SAFT).
%
SAFT is a theory for fluids in which hydrogen bonding plays a significant
role\cite{chapman1989saft, muller2001molecular}. SAFT is used to
accurately model the equations of state of both pure fluids and
mixtures over a wide range of temperatures and pressures.
\fixme{Talk about and cite papers using SAFT to build DFTs.}
%
In a recent paper \cite{hughes2013classical}, we constructed a
classical DFT for water which uses SAFT as a basis for a free energy
functional. We used Clark \emph{et al}'s \cite{clark2006developing}
five fitting parameters for an optimal cDFT for water and, in
addition, introduced a parameter for the length scale over which the
dispersion interaction is correlated.

SAFT is based on Wertheim's first-order thermodynamic perturbation
theory (TPT1)\cite{wertheim1984fluidsI, wertheim1984fluidsII,
  wertheim1986fluidsIII, wertheim1986fluidsIV}, which allows it to
account for strong associative interactions between
molecules. Wertheim's thermodynamic perturbation theory uses a
hard-sphere reference fluid where each sphere has a number of
association sites on the surface. The association sites have an
effective range and a small energy contribution when a site is within
range of another hard-sphere's site. The free energy due to this
association depends on the radial distribution function
$g_\sigma^\text{HS}(\rr)$. Our previous water functional used Yu and
Wu's radial distribution function
\cite{yu2002fmt-dft-inhomogeneous-associating}, which was found to
agree less with hard-sphere Monte-Carlo simulation than a distribution
function developed by us in a recent paper.\cite{schulte2012using} In
our paper, we developed an approximation for correlation at
contact to describe the hard-sphere fluid.

@@

We modify our previous the association free energy term to use our
$g_{\sigma}^A$ rather than Yu and Wu's $g_\sigma^{YuWu}$.  The
association free energy for our four-site model has the form:
\begin{align}
  F_\text{assoc}[n] &= k_BT \int n_\text{site}(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where
\begin{align}
  n_\text{site}(\rr) &=
  \begin{cases}
    4 n(\rr) & \text{this work}\\
    4 n_0(\rr) \zeta(\rr) & \text{Ref.~\citenum{hughes2013classical}}
  \end{cases} 
\end{align}
is the density of bonding sites at position~$\rr$.  The four in
$n_\text{site}(\rr)$ comes from the four hydrogen bond sites,
$n_0(\rr)$ is the average density contacting point $\rr$, and
$\zeta(\xx)$ is a dimensionless measure of the density inhomogeneity
from Yu and Wu~\cite{yu2002fmt-dft-inhomogeneous-associating}.  The
functional $X(\rr)$ is the fraction of association sites \emph{not}
hydrogen-bonded, which is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 2n_\text{site}'(\rr)
      \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)} - 1}
  {n_\text{site}'(\rr)
    \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)}, \label{eq:X}
\end{align}
where
\begin{align}
  n_\text{site}'(\rr) &=
  \begin{cases}
    \frac{4}{\pi\sigma^2} \int n(\rr')\delta(\sigma - |\rr-\rr'|) d\rr' & \text{this work}\\
    4 n_0(\rr) \zeta(\rr) & \text{Ref.~\citenum{hughes2013classical}}
  \end{cases} 
\end{align}
is the density of bonding sites that could bond to the sites~$n_\text{site}(\rr)$, and
\begin{align}
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW},
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well dispersion
potential, and $a_1$ and $a_2$ are the two terms in the dispersion
free energy.  The radial distribution function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid
correlation function $g^\textit{HS}_\sigma$. Four our old functional
we used Yu and Wu's radial distribution function, and for this paper we
use our $g^\textit{A}_\sigma$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/density-compare}
\end{center}
\caption{ Density profiles for a water around a single hydrophobic
  rod. The solid red profile is the density using our new functional
  and the dashed blue profile the result of our old functional.  For
  comparison, under the profiles is a cartoon of a string of hard
  spheres touching in one dimension. The horizontal black dotted line
  is the bulk density.}
\label{fig:density-single-rod}
\end{figure}

@@

We continue our study into three dimensions with a hard sphere solute
in water. Unlike the single rod, the number of hydrogen bonds does not appear to increase
linearly with the sphere radii we used. \fixme{When the size of
each rodis increase, the surface area increases proportional to radius.
However, when we increase the radius of each sphere, the surface area
increases as $R^2$.}

@@

We have modified a classical DFT for water with a more accurate radial
distribution function which changes the way that the molecules assosciate.

@@

A large fraction of interesting chemistry---including all of molecular
biology---takes place in aqueous solution.  However, while quantum
chemistry enables us to calculate the ground state energies of large
molecules in vacuum, prediction of the free energy of even the
smallest molecules in the presence of a solvent poses a continuing
challenge due to the complex structure of a liquid and the
computational cost of \emph{ab initio} molecular
dynamics~\cite{car1985, grossman2004}.  The current state-of-the art
in \emph{ab initio} molecular dynamics is limited to a few hundred
water molecules per unit cell~\cite{lewis2011doesnitric}.
%
On top of this, traditional density-functional theory (DFT) methods
without the use of dispersion corrections strongly over-structure
water, to the point that ice melts at over 120$^\circ$C\cite{yoo2009phase}!
%
There has been a flurry of recent
publications implicating van der Waals effects (i.e. dispersion
corrections) as significant in
reducing this over-structuring\cite{lin2009importance,
  wang2011density, mogelhoj2011ab, jonchiere2011van}.
%
However, one particular study found that water modeled using a
hybrid functional with dispersion corrections \emph{still} has a
melting point over 80$^\circ$C\cite{yoo2011}.
%
It has also
been found that the inclusion of nuclear quantum effects can provide
similar improvements \cite{morrone2008nuclear}.  Each of these
corrections imposes an additional computational burden on an approach
that is already feasible for only a very small number of water
molecules. A more efficient approach is needed in order to study
nanoscale and larger solutes.

@@

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, yang1994density, gloor2002saft,
  gloor2004accurate, gloor2007prediction, Jaqaman2004,
  clark2006developing, chuev2006, lischner2010classical,
  fu2005vapor-liquid-dft,kiselev2006new, blas2001examination,
  sundararaman2012computationally}, each of
which captures some of the qualitative behavior of water.  However,
each of these functionals also fail to capture some of water's unique
properties.  For instance, the functional of Lischner \emph{et
  al}\cite{lischner2010classical} treats the surface tension
correctly, but can only be used at room temperature, and thus captures
none of the temperature-dependence of water.  A functional by Chuev
and Skolov\cite{chuev2006} uses an ad hoc modification of FMT that can
predict hydrophobic hydration near temperatures of 298~K, but does not
produce a correct equation of state due to their method producing a
high value for pressure.
A number of classical density functionals
have recently been produced that are based on Statistical Associating
Fluid Theory (SAFT)\cite{ segura1997associating, segura1998comparison,
  yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft,gloor2002saft,muller2001molecular,
 clark2006developing, gloor2007prediction, gloor2004accurate,
  gross2009density, kahl2008modified, blas2001examination}.  These
functionals are based on a perturbative thermodynamic expansion, and
do reproduce the temperature-dependence of water's properties.
% 
We should give special mention to Sundararaman \emph{et al} who
recently introduced a classical density functional for water using a
model in which a water molecule is treated as a hard sphere attached to
two tetrahedrally oriented hard spheres representing voids, or
orientations in which a hydrogen bond may not be formed, with all
attractive interactions being lumped into a single pair potential
treated in a mean field
approximation\cite{sundararaman2012computationally}.

>>>>

\begin{abstract}
We develop an improved version of our old functional.
\end{abstract}
\maketitle

\section{Introduction}

Water, the universal solvent, is of critical practical importance, and
a continuum description of water is in high demand for a solvation
model.  A number of recent attempts to develop improved solvation
models for water have built on the approach of classical
density-functional theory (DFT)\cite{jeanmairet2013molecular,
  zhao2011molecular, zhao2011new, ramirez2005direct,
  ramirez2005density, levesque2012solvation, levesque2012scalar}
~\fixme{These are from Jeanmairet 2013 citations 35-44; only the ones
  that discuss water}.  Classical DFT is based on a description of a fluid written
as a free energy functional of the density distribution.  There are
two approaches used to construct a classical DFT for water.  The first
is to choose a functional form that is fit to properties of the bulk
liquid at a given temperature and pressure
\cite{jeanmairet2013molecular} ~\fixme{cite Jeanmairet and similar}.
Using this approach, it is possible to construct a functional that
reproduces the exact second-order response function of the liquid
under the fitted conditions.  This class of functional, however, will
be less accurate at other temperatures or pressures---and in the
inhomogeneous scenarios in which solvation models are applied.  The
second approach is construct a functional by applying liquid-state
theory to a model system, and then fit the model to experimental data
such as the equation of state.

\fixme{stuff about water}.  An important aspects of water
is its ability to form hydrogen bonds.  Each water molecule contains
two hydrogen bond donor sites and two acceptor sites.
%
One promising approach for an efficient continuum description of water
is that of classical density-functional theory (DFT), which is an
approach for evaluating the free energy and thermally averaged density
of fluids in an arbitrary external potential~\cite{ebner1976density}.
The foundation of classical DFT is the Mermin
theorem\cite{mermin1965thermal}, which extends the Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} to non-zero temperature.
Classical DFT is a natural framework for creating a more flexible
theory of hydrophobicity that can readily describe interaction of
water with arbitrary external potentials---such as potentials
describing strong interactions with solutes or surfaces.

A widely used family of models used in the development of classical
density functionals is based on Statistical Associating Fluid Theory
(SAFT).

SAFT is a theory for fluids in which hydrogen bonding plays a significant
role\cite{chapman1989saft, muller2001molecular}. SAFT is used to
accurately model the equations of state of both pure fluids and
mixtures over a wide range of temperatures and pressures.
\fixme{Talk about and cite papers using SAFT to build DFTs.}

In a recent paper \cite{hughes2013classical}, we constructed a
classical DFT for water which uses SAFT as a basis for a free energy
functional. We used Clark \emph{et al}'s \cite{clark2006developing}
five fitting parameters for an optimal cDFT for water and, in
addition, introduced a parameter for the length scale over which the
dispersion interaction is correlated.

SAFT is based on Wertheim's first-order thermodynamic perturbation
theory (TPT1)\cite{wertheim1984fluidsI, wertheim1984fluidsII,
  wertheim1986fluidsIII, wertheim1986fluidsIV}, which allows it to
account for strong associative interactions between
molecules. Wertheim's thermodynamic perturbation theory uses a
hard-sphere reference fluid where each sphere has a number of
association sites on the surface. The association sites have an
effective range and a small energy contribution when a site is within
range of another hard-sphere's site. The free energy due to this
association depends on the radial distribution function
$g_\sigma^\text{HS}(\rr)$. Our previous water functional used Yu and
Wu's radial distribution function
\cite{yu2002fmt-dft-inhomogeneous-associating}, which was found to
agree less with hard-sphere Monte-Carlo simulation than a distribution
function developed by us in a recent paper \cite{schulte2012using}. In
our paper, we developed an approximation for correlation at
contact to describe the hard-sphere fluid.

@@

We modify our previous association free energy term to use our
$g_{\sigma}^\textit{A}$ rather than Yu and Wu's $g_\sigma$.  The
association free energy for our four-site model has the form:
\begin{align}
  F_\text{assoc}[n] &= k_BT \int n_\text{site}(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where
\begin{align}
  n_\text{site}(\rr) &=
  \begin{cases}
    4 n(\rr) & \text{this work}\\
    4 n_0(\rr) \zeta(\rr) & \text{Ref.~\citenum{hughes2013classical}}
  \end{cases} 
\end{align}
is the density of bonding sites at position~$\rr$.  The four in
$n_\text{site}(\rr)$ comes from the four hydrogen bond sites,
$n_0(\rr)$ is the average density contacting point $\rr$, and
$\zeta(\xx)$ is a dimensionless measure of the density inhomogeneity
from Yu and Wu~\cite{yu2002fmt-dft-inhomogeneous-associating}.  The
functional $X(\rr)$ is the fraction of association sites \emph{not}
hydrogen-bonded, which is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 2n_\text{site}'(\rr)
      \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)} - 1}
  {n_\text{site}'(\rr)
    \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)}, \label{eq:X}
\end{align}
where
\begin{align}
  n_\text{site}'(\rr) &=
  \begin{cases}
    \frac{4}{\pi\sigma^2} \int n(\rr')\delta(\sigma - |\rr-\rr'|) d\rr' & \text{this work}\\
    4 n_0(\rr) \zeta(\rr) & \text{Ref.~\citenum{hughes2013classical}}
  \end{cases} 
\end{align}
is the density of bonding sites that could bond to the sites~$n_\text{site}(\rr)$, and
\begin{align}
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW},
\end{align}
This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well dispersion
potential, and $a_1$ and $a_2$ are the two terms in the dispersion
free energy.  The radial distribution function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid
correlation function $g^\textit{HS}_\sigma$. Four our old functional
we used Yu and Wu's radial distribution function, and for this paper we
use our $g^\textit{A}_\sigma$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/density-compare}
\end{center}
\caption{ Density profiles for a water around a single hydrophobic
  rod of radius 0.2 nm. The solid red profile is the density using our new functional
  and the dashed blue profile the result of our old functional.  For
  comparison, under the profiles is a cartoon of a string of hard
  spheres touching in one dimension. The horizontal black dotted line
  is the bulk density for water.}
\label{fig:density-single-rod}
\end{figure}

@@

We continue our study in three dimensions with a hard sphere solute in
water. Similar to the single rod, we see a decrease in the number of
broken hydrogen bonds for a given rod radius. However, with the single
rods the number of hydrogen bonds appears to increase linearly with
rod radius where the sphere radii we used. \fixme{When the size of
  each rod is increased, the surface area increases proportional to
  radius.  However, when we increase the radius of each sphere, the
  surface area increases as $R^2$.}

@@

We have modified a classical DFT for water with a more accurate radial
distribution function which modifies assosciation between water
molecules. This modification predicts less broken hydrogen bonds due
to a hard rod, and solutes placed in water than the previous theory.

@@

A large fraction of interesting chemistry---including all of molecular
biology---takes place in aqueous solution.  However, while quantum
chemistry enables us to calculate the ground state energies of large
molecules in vacuum, prediction of the free energy of even the
smallest molecules in the presence of a solvent poses a continuing
challenge due to the complex structure of a liquid and the
computational cost of \emph{ab initio} molecular
dynamics~\cite{car1985, grossman2004}.  The current state-of-the art
in \emph{ab initio} molecular dynamics is limited to a few hundred
water molecules per unit cell~\cite{lewis2011doesnitric}.

On top of this, traditional density-functional theory (DFT) methods
without the use of dispersion corrections strongly over-structure
water, to the point that ice melts at over 120$^\circ$C\cite{yoo2009phase}!

There has been a flurry of recent
publications implicating van der Waals effects (i.e. dispersion
corrections) as significant in
reducing this over-structuring\cite{lin2009importance,
  wang2011density, mogelhoj2011ab, jonchiere2011van}.

However, one particular study found that water modeled using a
hybrid functional with dispersion corrections \emph{still} has a
melting point over 80$^\circ$C\cite{yoo2011}.

It has also
been found that the inclusion of nuclear quantum effects can provide
similar improvements \cite{morrone2008nuclear}.  Each of these
corrections imposes an additional computational burden on an approach
that is already feasible for only a very small number of water
molecules. A more efficient approach is needed in order to study
nanoscale and larger solutes.

@@

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, yang1994density, gloor2002saft,
  gloor2004accurate, gloor2007prediction, Jaqaman2004,
  clark2006developing, chuev2006, lischner2010classical,
  fu2005vapor-liquid-dft,kiselev2006new, blas2001examination,
  sundararaman2012computationally}, each of
which captures some of the qualitative behavior of water.  However,
each of these functionals also fail to capture some of water's unique
properties.  For instance, the functional of Lischner \emph{et
  al}\cite{lischner2010classical} treats the surface tension
correctly, but can only be used at room temperature, and thus captures
none of the temperature-dependence of water.  A functional by Chuev
and Skolov\cite{chuev2006} uses an ad hoc modification of FMT that can
predict hydrophobic hydration near temperatures of 298~K, but does not
produce a correct equation of state due to their method producing a
high value for pressure.
A number of classical density functionals
have recently been produced that are based on Statistical Associating
Fluid Theory (SAFT)\cite{ segura1997associating, segura1998comparison,
  yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft,gloor2002saft,muller2001molecular,
 clark2006developing, gloor2007prediction, gloor2004accurate,
  gross2009density, kahl2008modified, blas2001examination}.  These
functionals are based on a perturbative thermodynamic expansion, and
do reproduce the temperature-dependence of water's properties.

We should give special mention to Sundararaman \emph{et al} who
recently introduced a classical density functional for water using a
model in which a water molecule is treated as a hard sphere attached to
two tetrahedrally oriented hard spheres representing voids, or
orientations in which a hydrogen bond may not be formed, with all
attractive interactions being lumped into a single pair potential
treated in a mean field
approximation\cite{sundararaman2012computationally}.
